---
title: "Power analysis"
author: "James Galante"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Save Image for Debugging

```{r}
# Saving image for debugging
save.image("RDA_objects/analyze_duplicate_k562_dc_tap_pairs.rda")
message("Saved Image")
stop("Manually Stopped Program after Saving Image")
```

### Goal

Exploration of duplicate EG pairs between K562 DC TAP and training dataset. Also, looking at the per guide effect size for the K562 DC TAP dataset. 

```{r requiredPackages, warning=FALSE, message=FALSE}
library(tidyverse)
library(GenomicRanges)
library(cowplot)
```

------------------------------------------------------------------------

```{r, message = FALSE}
# Load in the data
# combined_unfiltered_k562_dc_tap <- read_tsv(snakemake@input$combined_unfiltered_k562_dc_tap) # ValidConnection Only
combined_unfiltered_k562_dc_tap <- read_tsv("/oak/stanford/groups/engreitz/Users/jgalante/DC_TAP_Paper/results/analyze_validation_datasets/dc_tap_plots/expt_pred_merged_annot/labelled_combined_validation_expt_pred_merged_annot.txt")
combined_unfiltered_k562_dc_tap <- combined_unfiltered_k562_dc_tap %>% filter(category == "K562 DC TAP Seq")


combined_training <- read_tsv(snakemake@input$combined_training)
gasperini_MAST_and_Sceptre <- readRDS(snakemake@input$gasperini_MAST_and_Sceptre)

# K562 DC TAP outlier / guide analysis
per_guide_effect_sizes_unfiltered_k562_dc_tap <- read_tsv(snakemake@input$per_guide_effect_sizes_unfiltered_k562_dc_tap)
grna_target_table <- read_tsv(snakemake@input$grna_target_table)
bonferroni_corrected_k562_dc_tap <- readRDS(snakemake@input$bonferroni_corrected_k562_dc_tap)

# k562_dc_tap_discovery_results <- read_tsv(snakemake@input$k562_dc_tap_discovery_results) # All Results
k562_dc_tap_discovery_results <- readRDS("/oak/stanford/groups/engreitz/Users/jgalante/DC_TAP_Paper/results/process_validation_datasets/K562_DC_TAP_Seq/differential_expression/results_run_discovery_analysis.rds")
k562_dc_tap_discovery_results %>% filter(!str_detect(grna_target, "_TSS"))


k562_dc_tap_gene_mapping <- read_tsv(snakemake@input$k562_dc_tap_gene_mapping, col_names = FALSE)
```

```{r, message = FALSE}
# Create the GRanges object
gr_combined_training <- GRanges(seqnames = combined_training$chrom, 
                                ranges = IRanges(start = combined_training$chromStart, 
                                                 end = combined_training$chromEnd), 
                                measuredGeneSymbol = combined_training$measuredGeneSymbol, 
                                training = combined_training)

gr_combined_unfiltered_k562_dc_tap <- GRanges(seqnames = combined_unfiltered_k562_dc_tap$chrom,
                                              ranges = IRanges(start = combined_unfiltered_k562_dc_tap$chromStart,
                                                               end = combined_unfiltered_k562_dc_tap$chromEnd),
                                              measuredGeneSymbol = combined_unfiltered_k562_dc_tap$measuredGeneSymbol,
                                              dc_tap = combined_unfiltered_k562_dc_tap)

# Find the overlaps and subset
overlaps <- findOverlaps(gr_combined_unfiltered_k562_dc_tap, gr_combined_training)
dc_tap_overlaps <- gr_combined_unfiltered_k562_dc_tap[queryHits(overlaps)]
training_overlaps <- gr_combined_training[subjectHits(overlaps)]

# Only keep the pairs overlaps that have the same measuredGeneSymbol
matches <- dc_tap_overlaps$measuredGeneSymbol == training_overlaps$measuredGeneSymbol
dc_tap_duplicates <- dc_tap_overlaps[matches] %>% as.data.frame() %>% mutate(id = seq_along(1:sum(matches)))
training_duplicates <- training_overlaps[matches] %>% as.data.frame() %>% mutate(id = seq_along(1:sum(matches)))

# Merge the two dataframes by ID (added in lines above)
duplicates <- dc_tap_duplicates %>%
  left_join(training_duplicates, by = "id", suffix = c(".dc_tap", ".training"))
```

## Duplicates b/t K562 DC TAP Seq and Full Training Data
```{r, message = FALSE}
duplicates %>%
  ggplot(aes(x = training.EffectSize * 100, y = dc_tap.EffectSize * 100, color = training.Reference)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Comparing Duplicate Pairs b/t DC TAP Seq and Training Data",
    y = "DC TAP Seq Dataset Effect Sizes (%)",
    x = "Training Dataset Effect Sizes (%)",
    color = "Training Dataset"
  ) +
  theme_bw() +
  ylim(-100, 20) +
  xlim(-100, 20) +
  scale_color_brewer(palette = "Set1") +
  theme(
    aspect.ratio = 1,
    panel.grid.minor = element_blank()
  )
```

## Let's Isolate Gasperini w & w/o Outlier
```{r, message = FALSE}
p1 <- duplicates %>%
  filter(training.Dataset == "Gasperini2019") %>%
  ggplot(aes(x = training.EffectSize * 100, y = dc_tap.EffectSize * 100)) +
  geom_point(color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  labs(
    title = "Comparing Duplicate Pairs",
    y = "DC TAP Seq Dataset Effect Sizes (%)",
    x = "Gasperini Dataset Effect Sizes (%)",
    subtitle = "DC TAP (Sceptre), Gasperini (MAST) [Zoomed]\nAll Data"
  ) +
  theme_bw() +
  ylim(-50, 10) +
  xlim(-50, 10) +
  scale_color_brewer(palette = "Set1") +
  theme(
    aspect.ratio = 1,
    panel.grid.minor = element_blank()
  )

# There is one outlier where the DC TAP Effect Size is near -3%, but Gasperini is near -35%. When we remove this outlier, it seems that the correlation is great. 

# print(duplicates %>% filter(measuredGeneSymbol.dc_tap == "RTN4"))

p2 <- duplicates %>%
  filter(training.Dataset == "Gasperini2019") %>%
  filter(measuredGeneSymbol.dc_tap != "RTN4") %>%
  ggplot(aes(x = training.EffectSize * 100, y = dc_tap.EffectSize * 100)) +
  geom_point(color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  labs(
    title = "Comparing Duplicate Pairs",
    y = "DC TAP Seq Dataset Effect Sizes (%)",
    x = "Gasperini Dataset Effect Sizes (%)",
    subtitle = "DC TAP (Sceptre), Gasperini (MAST) [Zoomed]\nWithout Outlier"
  ) +
  theme_bw() +
  ylim(-50, 10) +
  xlim(-50, 10) +
  scale_color_brewer(palette = "Set1") +
  theme(
    aspect.ratio = 1,
    panel.grid.minor = element_blank()
  )

plot_grid(p1, p2)
```


## Gasperini MAST versus Sceptre

We See that the effect sizes correlation between DC TAP Effect sizes and Gasperini Effect sizes without that one outlier is very good, but we should compare these effect sizes with similar methodologies.

```{r, message = FALSE}
# Create a color column based on the classification of significance b/t MAST and Sceptre
gasperini_MAST_and_Sceptre <- gasperini_MAST_and_Sceptre %>%
  mutate(
    color_category = case_when(
      significant == TRUE & Sceptre_pctChange < 0 & Regulated == TRUE ~ "purple",   # Sceptre and MAST
      significant == TRUE & Sceptre_pctChange < 0 & Regulated == FALSE ~ "blue",  # Sceptre Only
      Regulated == TRUE & significant == FALSE ~ "red",                            # MAST Only
      TRUE ~ "grey"                                                                 # Neither
    )
  )
```

```{r, message = FALSE}
# Separate grey points and colored points to control point z axis
grey_points <- gasperini_MAST_and_Sceptre %>% filter(color_category == "grey")
colored_points <- gasperini_MAST_and_Sceptre %>% filter(color_category != "grey")

gasperini_MAST_and_Sceptre %>%
  ggplot() +
  geom_point(data = grey_points, aes(x = MAST_pctChange, y = Sceptre_pctChange), color = "grey", alpha = 1, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red", size = 0.7, aes(linetype = "y ~ x")) +
  geom_point(data = colored_points, aes(x = MAST_pctChange, y = Sceptre_pctChange, color = color_category),
             alpha = 1, size = 0.5) +
  geom_smooth(data = gasperini_MAST_and_Sceptre, aes(x = MAST_pctChange, y = Sceptre_pctChange), method = "lm",
              se = TRUE, color = "black", linetype = "solid") +
  geom_smooth(data = colored_points %>% filter(color_category == "purple"), 
              aes(x = MAST_pctChange, y = Sceptre_pctChange), method = "lm",
              se = TRUE, color = "purple", linetype = "solid") +
  labs(
    title = "Correlation b/t MAST and Sceptre Results for Gasperini",
    x = "MAST Effect Size (%)",
    y = "Sceptre Effect Size (%)",
    color = "Significance"
  ) +
  # Define manual colors and labels
  scale_color_manual(
    values = c("purple" = "purple", "blue" = "blue", "red" = "red"),
    labels = c(
      "purple" = "Sceptre and MAST",
      "blue" = "Sceptre Only",
      "red" = "MAST Only"
    )
  ) +
  # Set axis limits
  coord_cartesian(xlim = c(-80, 45), ylim = c(-80, 45)) +
  # Apply theme
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.title = element_text(face = "bold"),
    aspect.ratio = 1
  )
```

It looks like MAST systematically underestimates the effect sizes when compared to Sceptre. This might mean that our "duplicates" analysis results could be misleading, but let's see.

If we expect Sceptre effect sizes to be larger, we might see a "shift" to the left of our points.

## Let's now see how the duplicates look with Sceptre Effect Sizes
```{r, message = FALSE}
# Merge the dataframes based on name
duplicates_w_sceptre <- gasperini_MAST_and_Sceptre %>%
  select(name, Sceptre_pctChange, MAST_pctChange, color_category) %>%
  right_join(duplicates %>% filter(training.Dataset == "Gasperini2019"), by = c("name" = "training.name"))
```

```{r, message = FALSE}
p1 <- duplicates_w_sceptre %>%
  ggplot(aes(x = Sceptre_pctChange, y = dc_tap.EffectSize * 100)) +
  geom_point(color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  labs(
    title = "Comparing Duplicate Pairs",
    y = "DC TAP Seq Dataset Effect Sizes (%)",
    x = "Gasperini Dataset Effect Sizes (%)",
    subtitle = "DC TAP (Sceptre), Gasperini (Sceptre) [Zoomed]\nAll Data"
  ) +
  theme_bw() +
  ylim(-50, 10) +
  xlim(-50, 10) +
  scale_color_brewer(palette = "Set1") +
  theme(
    aspect.ratio = 1,
    panel.grid.minor = element_blank()
  )

### How does the previous outlier look here
p2 <- duplicates_w_sceptre %>%
  filter(measuredGeneSymbol.dc_tap != "RTN4") %>%
  ggplot(aes(x = Sceptre_pctChange, y = dc_tap.EffectSize * 100)) +
  geom_point(color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  labs(
    title = "Comparing Duplicate Pairs",
    y = "DC TAP Seq Dataset Effect Sizes (%)",
    x = "Gasperini Dataset Effect Sizes (%)",
    subtitle = "DC TAP (Sceptre), Gasperini (Sceptre) [Zoomed]\nWithout Outlier"
  ) +
  theme_bw() +
  ylim(-50, 10) +
  xlim(-50, 10) +
  scale_color_brewer(palette = "Set1") +
  theme(
    aspect.ratio = 1,
    panel.grid.minor = element_blank()
  )

plot_grid(p1, p2)
```

So there still seems to be a strong correlation, but we do indeed see a shift to the left. This could also mean that the DC TAP Seq data is in fact underestimating effect sizes

## Let's look into this outlier
```{r}
# Create a color column based on the classification of significance b/t duplicates
duplicates_w_sceptre <- duplicates_w_sceptre %>%
  mutate(
    color_category = case_when(
      color_category == "purple" & dc_tap.Regulated == TRUE ~ "purple",   # DC TAP & Gasperini
      color_category == "grey" & dc_tap.Regulated == TRUE ~ "blue",       # DC TAP Only 
      color_category == "purple" & dc_tap.Regulated == FALSE ~ "red",     # Gasperini Only
      TRUE ~ "grey"                                                                 
    )
  )

duplicates_w_sceptre %>%
  ggplot(aes(x = Sceptre_pctChange, y = dc_tap.EffectSize * 100, color = color_category)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  labs(
    title = "Comparing Duplicate Pairs",
    y = "DC TAP Seq Dataset Effect Sizes (%)",
    x = "Gasperini Dataset Effect Sizes (%)",
    color = "Significance",
    subtitle = "DC TAP (Sceptre), Gasperini (Sceptre) [Zoomed]\nAll Data"
  ) +
  theme_bw() +
  ylim(-50, 10) +
  xlim(-50, 10) +
  scale_color_manual(
    values = c("purple" = "purple", "blue" = "blue", "red" = "red"),
    labels = c(
      "purple" = "DC TAP & Gasperini",
      "blue" = "DC TAP Only",
      "red" = "Gasperini Only")
  ) +
  theme(
    aspect.ratio = 1,
    panel.grid.minor = element_blank()
  )
```

It looks like the outlier is significant in both datasets, so why is the sceptre effect size so low?
It's also of note that a majority of the points in this correlation are insignificant in either Sceptre or MAST.

# Outlier Analysis
## Let's check the p value histogram for the per guide significance testing
```{r}
per_guide_effect_sizes_unfiltered_k562_dc_tap %>%
  ggplot(aes(x = p_value)) +
  geom_histogram(bins = 300) +
  labs(
    title = "Histogram of p_values for per guide significance testing",
    subtitle = "300 bins",
    x = "P-Value",
    y = "Count"
  ) +
  theme_classic()
```

The histogram looks great albeit some plotting gaps, which are fine.

# The outlier
The dc tap name for this outlier is `RTN4|chr2:55368591-55369057:.`
The gasperini name for this outlier is `RTN4|chr2:55367486-55368618:.`

Looking at the IGV session for these guides, we find that they are very different in terms of accessibility even though they're under the same ENCODE pert



